Supplementary Information

### 2021-01-12 doc/manuscript/MEE_manuscript.Rmd
finish filling in results

```{r , eval=FALSE, echo=FALSE}
## Tallies for the  paragraph below (Jan12)

@@

###Windows selected: 
# BCNM: precip dry season (12mo), T dry season (2 mo) 
#HKK: PDF dry season + edges (9mo), T wet season (8 mo)
#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo)
#LDW: precip c June (1 mo), PET current June (1mo)
#HF: PDF pMay (1mo), PET cMar (1mo)
#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo)
#NIO: wet cMay (1mo), TMP pJune (1mo)
#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo)
#CB: P pJune-cJul(14mo), TMXpAug-pAug(1mo)
#SC: P pJun-pAug(3mo), TMXcJul (1mo)
```
Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but one species, and were predominantly positive (Fig. `r clim_sens`).
$PPT$ and $PDF$ were each identified as the top variable in four of the eight sites with both variables available, while $PPT$ was the only option at two (LT and CB).
Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  wet seasons in the tropics (BCNM and HKK) or late spring/ early summer in the extratropics (n= 5 of 8).
In the tropics, the long time windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
At two extratropical sites (LDW, NIO), the most influential windows were limited to the current spring and early summer, whereas optimal windows at three sites extended back to the previous fall (LT, CB) or summer (SCBI).
Finally, at three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but two species (*Tsuga canadensis* at HF and *Betula papyrifera* at NIO; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (*# of #* with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., BEAL at HF; Fig. `r clim_sens`). 
At the riparian site (NIO), for which streamflow was included as a candidate variable in the precipitation group, May $PDF$ was selected as the strongest driver but had a modest negative effect on growth, indicating that water availability was not a significant limitation for *Betula papyrifera* at this site.

A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses (*#*% of species; Fig. `r clim_sens`).
The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two sites. 
$T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCNM, respectively), noting that the effects of $T_{min}$ at BCNM were only marginally significant (Fig. `r clim_sens`).
Optimal time windows most commonly occurred during the peak growing season of the current year (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (*i.e.*, negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean). 
In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March $PET$ for *Tsuga Canadensis* at HF, the latter contrasting with a negative response of the other three species at HF (Fig. `r clim_sens`).
At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig. `r SC_1970break`).

>>>>

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Jan12)

@@

###Windows selected: 
# BCNM: precip dry season (12mo), T dry season (2 mo) 
#HKK: PDF dry season + edges (9mo), T wet season (8 mo)
#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo)
#LDW: precip c June (1 mo), PET current June (1mo)
#HF: PDF pMay (1mo), PET cMar (1mo)
#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo)
#NIO: wet cMay (1mo), TMP pJune (1mo)
#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo)
#CB: P pJune-cJul(14mo), TMXpAug-pAug(1mo)
#SC: P pJun-pAug(3mo), TMXcJul (1mo)

###other:
n_pos_first_order_P_term=42 #(exceptions: BEPA & TSCA)
n_pos_1st_P_term_and_neg_2nd = 32
percent_temperature_responses_1st_order_negative=round(13/38*100)
percent_temperature_responses_1st_order_pos_2nd_order_neg=19/38*100
```
Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but one species, and were predominantly positive (Fig. `r clim_sens`).
$PPT$ and $PDF$ were each identified as the top variable in four of the eight sites with both variables available, while $PPT$ was the only option at two (LT and CB).
Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  wet seasons in the tropics (BCNM and HKK) or late spring/ early summer in the extratropics (n= 5 of 8).
In the tropics, the long time windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
At two extratropical sites (LDW, NIO), the most influential windows were limited to the current spring and early summer, whereas optimal windows at three sites extended back to the previous fall (LT, CB) or summer (SCBI).
Finally, at three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but two species (*Tsuga canadensis* at HF and *Betula papyrifera* at NIO; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., BEAL at HF; Fig. `r clim_sens`). 
At the riparian site (NIO), for which streamflow was included as a candidate variable in the precipitation group, May $PDF$ was selected as the strongest driver but had a modest negative effect on growth, indicating that water availability was not a significant limitation for *Betula papyrifera* at this site.

A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses, particularly at the higher end of the temperature range (`r percent_temperature_responses_1st_order_negative + percent_temperature_responses_1st_order_pos_2nd_order_neg`%; `r percent_temperature_responses_1st_order_negative`% with negative first-order term, `r percent_temperature_responses_1st_order_pos_2nd_order_neg`% with positive first-order term but negative second-order term; Fig. `r clim_sens`).
The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two sites. 
$T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCNM, respectively), noting that the effects of $T_{min}$ at BCNM were only marginally significant (Fig. `r clim_sens`).
Optimal time windows most commonly occurred during the peak growing season of the current year (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (*i.e.*, negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean). 
In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March $PET$ for *Tsuga Canadensis* at HF, the latter contrasting with a negative response of the other three species at HF (Fig. `r clim_sens`).
At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig. `r SC_1970break`).
